---
title: "trade-offs_meta-analysis_2026.Rmd"
author: "maciejB"
date: "2026-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis is related to the study:

"Growth–reproduction trade-offs are common but changing in woody plants: a meta-analysis" 
Barczyk et al. 2026

Summary
Growth and reproduction draw on a common resource pool, yet empirical studies of woody plants report widely differing relationships between seed production and growth. Here we synthesize 685 estimates from 78 studies covering 79 woody species, with study locations concentrated at northern temperate latitudes, to test how growth–reproduction correlations vary across time, species, and environments. Growth and reproduction measured within the same year were negatively correlated, suggesting an immediate cost of reproduction. The strength and direction of growth–reproduction correlations showed at most weak phylogenetic structure among the sampled species and were not systematically related to functional traits and climate. Instead, trade-offs were strongest in species with high interannual variability in seed production and weakened markedly over recent decades in these species. Together, these results show that growth–reproduction trade-offs in woody plants are common but not fixed, and that shifts in reproductive variability under environmental change can alter how trees balance growth and reproduction, with consequences for long-term forest functioning.


```{r setup, include=FALSE}
 
usethis::git_sitrep()
usethis::use_git()

```

```{r packages}

library(here)
library(metafor)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(DT)
library(meta)
library(lme4)
library(lmerTest)
library(mgcv)
library(hrbrthemes)
library(tidyr)
library(viridis)
library(RColorBrewer)
library(gplots)
library(rmarkdown)
library(knitr)
library(ape)
library(phytools); library(readxl); library(readr) ;library(ggtree);
library(patchwork); library(ggeffects); library(ggthemes)
library(glmmTMB); library(emmeans) ; library(reshape2)
library(rnaturalearth); library(sf) ; library(splines)
library(geosphere); library(ncf); library(raster)
library(wesanderson)
 
#install.packages("pacman")
pacman::p_load(devtools,R.rsp)

#devtools::install_github("daniel1noble/orchaRd", force = TRUE)
library(orchaRd)

# devtools::install_github("jinyizju/V.PhyloMaker2")
library(V.PhyloMaker2)

#devtools::install_github("thomasp85/patchwork")
library(patchwork)

### install package with whittaker plot from github
# devtools::install_github("valentinitnelav/plotbiomes") 
library(plotbiomes)

```

```{r data loading, echo=FALSE}

here()
 
### read the final meta-analysis data frame
meta <- readxl::read_xlsx(here("DATA_files","trade-offs_meta-analysis-data.xlsx")) %>% filter(decision != "NA")
str(meta)

## prepare the data frame converting variables to a proper format
meta <- meta %>%
  mutate(across(c(longitude, latitude, number_of_trees, number_of_years, `sample_size_ifY=1`), as.numeric),
         totalN = number_of_trees*number_of_years,
         tip_label = str_replace(species,' ','_'),
        plant_type = str_to_title(plant_type))
         
## count full number of studies
meta   %>%  group_by(title) %>% tally()

# 146 studies included after abstract screening

## prepare the final data frame with only valid observations
meta_INC <-  meta %>%
  filter(decision == "INCLUDE" & Z != "NA")

str(meta_INC)      

## count n of observations by title and population_id
meta_INC %>%  group_by(title) %>% tally() 
meta_INC %>%  group_by(pop_id) %>% tally() 

## count number of studies with correlations restricted to the same year
meta_INC %>%
  filter(H1 =="zero") %>%   group_by(title) %>% tally() 

meta_INC %>%
  filter(H1 =="zero") %>%     group_by(pop_id) %>% tally() 

# 738 effect sizes
# 78 studies
# 79 species (tip_label)
# 308 populations
# 38 genera

meta   %>%  filter(Rayyan =="present") %>%  group_by(title) %>% tally() 
meta_INC %>% filter(Rayyan =="present") %>%  group_by(title) %>% tally() 

# 48 eligible studies found by Rayyan (from 110 studies identified)

## count records with extra information
meta %>%
  filter(decision == "extra information") 
meta_INC %>% filter(H1 == "NA")
# 51 effect sizes with extra lags and 19 effect sizes with extra information = 70

```

PHYLOGENY 
the leading phylogeny in the analysis is the phylogeny from Zanne et al. 2014
the phylogeny from the open tree of life (synthetic tree lacking the divergence time of clades) 
but up to date list of species will be used in the supplement for the sensitivity analysis

construct the phylogenetic tree for species in the meta-analysis

```{r}
 
### summarizes all species names from meta-analysis
meta %>% 
group_by(species) %>%
summarize(tip_label = unique(tip_label)) 

### summarizes species names from meta-analysis (final data with taxa names)
my_taxa <- meta_INC  %>% 
group_by(tip_label) %>% 
summarize(tip_label = unique(tip_label)) 

### upload phylogenetic tree (Zanne et al.2016) and change 0 to minimal values

phylogenetictree  <-  ape::read.tree(file=here("DATA_files","Vascular_Plants_rooted.dated.tre")) 
phylogenetictree$edge.length[which(phylogenetictree$edge.length == 0)] = 1e-10

### compare misssing species from the final set of species compared to the uploaded phylogenetic tree
setdiff(my_taxa$tip_label, phylogenetictree$tip.label)
  
## find similar names from rotl package
### adds missing taxa

miss_taxa <- c("Larix_principis-rupprechtii" ,"Liquidambar_styracifula", "Ocotea_tenera"  , "Quercus_coccinea" , "Quercus_douglasii", "Quercus_liaotungensis","Quercus_phellos",  "Quercus_prinus" ,  "Rhamnus_alpinus")

### compare missing names with the names used in Open Tree of Life (rotl)
my_taxa_names <- as.character(my_taxa$tip_label)

resolved_names <- rotl::tnrs_match_names(names = miss_taxa, context_name = "All life")
class(resolved_names)
resolved_names$unique_name

```
## see all Quercus species present in the tree

all_quercus_tips <- grep(
  "^Quercus[ _]",
  phylogenetictree$tip.label,
  value = TRUE
)

all_quercus_tree <- ape::keep.tip(
  phylogenetictree,
  tip = all_quercus_tips
)

ggtree(all_quercus_tree) +
  geom_tiplab(fontface = "italic") +
  geom_treescale()
 


```{r  changing names to keep the phylogenetic tree updated}

## new names to adjust to the phylogenetic tree (synonyms found in the internet or rotl)

newNames <- tibble(
  old = c("Larix_principis-rupprechtii", "Quercus_liaotungensis", "Rhamnus_alpinus", "Quercus_prinus", "Liquidambar_styracifula"),
  new = c("Larix_gmelinii", "Quercus_mongolica", "Rhamnus_alpina","Quercus_montana","Liquidambar_styraciflua")
)

## replacing names across data frames

my_taxa <- my_taxa %>%
  left_join(newNames, by = c("tip_label" = "old")) %>%
  mutate(tip_label = coalesce(new, tip_label)) %>%
  dplyr::select(-new)
  
## replacing original names with the names in phylogenetic tree
meta <- meta %>%
  left_join(newNames, by = c("tip_label" = "old")) %>%
  mutate(tip_label = coalesce(new, tip_label)) %>%
dplyr::select(-new)
  
  meta_INC <- meta_INC %>%
  left_join(newNames, by = c("tip_label" = "old")) %>%
  mutate(tip_label = coalesce(new, tip_label)) %>%
dplyr::select(-new)
  
### compare species from two data sets again (the corrected species list and available phylogenetic tree)
setdiff(my_taxa$tip_label, phylogenetictree$tip.label)

phylolabel <- phylogenetictree$tip.label

######## adding missing species to the phylogenetic tree

# Ocotea tenera 

## Ocotea oblonga is the closest relative present in the tree according to Penagos et al. 2020 
 
new_phylo_updated1 <- phytools::bind.tip(
  phylogenetictree,
  tip.label = "Ocotea_tenera",
  where = which(phylogenetictree$tip.label == "Ocotea_oblonga"),
  position = 1.9,
  edge.length = 1.9
)

## plot tree for Ocotea genus to see the edits

tree_oc <- ape::keep.tip(
  new_phylo_updated1,
  tip = grep(
  "^Ocotea[ _]",
  new_phylo_updated1$tip.label,
  value = TRUE
) )

ggtree(tree_oc, layout = "fan") +
  geom_tiplab(fontface = "italic") +
  geom_treescale()
 

# Quercus phylogeny according to Zhou et al. 2022 (NatCom) and Althaus et al. 2026 (PNAS)

### Quercus douglasii

new_phylo_updated2 <- bind.tip(new_phylo_updated1,
                tip.label   = "Quercus_douglasii",
                where       = which(phylogenetictree$tip.label == "Quercus_lobata"),  # attach at that node
                position    = 0.01,          # 0 => attach at the node (polytomy)
                edge.length = 0.01)       # choose a branch length (units match your tree)


### Quercus coccinea (section Lobatae)

new_phylo_updated3 <- phytools::bind.tip(
  new_phylo_updated2,
  tip.label = "Quercus_coccinea",
  where = which(phylogenetictree$tip.label == "Quercus_velutina"),
  position = 0.01,
  edge.length = 0.01
)

### Quercus phellos

new_phylo_updated    <- bind.tip(new_phylo_updated3,
                tip.label   = "Quercus_phellos",
                where       =  which(phylogenetictree$tip.label == "Quercus_laurifolia"),  
                position    = 0.3,          # 0 => attach at the node (polytomy)
                edge.length = 0.3)       # choose a branch length (units match your tree)
                

### ####
####
###
## plot tree for Quercus genus to see the edits

tree_quercus <- ape::keep.tip(
  new_phylo_updated,
  tip = grep(
"^Quercus[ _]",
  new_phylo_updated$tip.label,
  value = TRUE
) )
 
 
ggtree(tree_quercus, layout = "fan", as.Date = TRUE, yscale_mapping = FALSE, hang = 0.1) +
  geom_tiplab(fontface = "italic") +
  geom_treescale()
                
############################################################################
### save the update phylogenetic tree
                        
write.tree(new_phylo_updated, file = "phylogenetic_tree_meta-updated.newick")

### subset tree withs species present in the data (original file and missing names)
subtree0  <-  drop.tip(phylogenetictree, setdiff(phylogenetictree$tip.label, meta_INC$tip_label))

subtree  <-  drop.tip(new_phylo_updated, setdiff(new_phylo_updated$tip.label, my_taxa$tip_label))

plot(subtree0)

# Plot the tree, and add the time axis
ape::plot.phylo(subtree, cex = 1) 
#ape::axisPhylo()

## plot the ciruclar tree

ggtree(subtree,layout = "fan", open.angle = 0) +
  geom_tiplab(fontface = "italic") +
  geom_treescale() + theme_tree2()

ggtree(subtree0,layout = "fan") +
  geom_tiplab(fontface = "italic") 

## IMPORTANT STEP FOR FURTHER ANALYSIS
## creates correlation matrix of phylogenetic relatedness 

phylo_corMatrix <- vcv(subtree, cor = T)

```





```{r construction of phylogenetic tree from open tree of life}

## get names from open tree of life
rotl::tol_about()

resolved_names <- rotl::tnrs_match_names(names = my_taxa$tip_label, context_name = "All life")

my_tree <- rotl::tol_induced_subtree(ott_ids = resolved_names$ott_id, label_format = "name")
  
##
is.binary(my_tree) # Check if tree is binary 
set.seed(1) # Set a seed to resolve politomies at random, and obtain similar results
binary.tree <- multi2di(my_tree, random=T) #resolve polytomies at random

my_phylo_branched <- compute.brlen(binary.tree, method = "Grafen", power = 1) # compute branch lengths using Grafen's method

### change names to adjust to my data

setdiff(my_taxa$tip_label, my_phylo_branched$tip.label)

my_phylo_branched$tip.label 

# change names to adjust data
# Oreoherzogia replace with Rhamnus
my_phylo_branched$tip.label  <- gsub("Oreoherzogia_alpina", "Rhamnus_alpina", my_phylo_branched$tip.label)

# Fuscospora replace with Nothofagus
my_phylo_branched$tip.label <- gsub("Fuscospora_truncata", "Nothofagus_truncata", my_phylo_branched$tip.label)  

# edit strange appendix but check its position first ()
my_phylo_branched$tip.label[55]
my_phylo_branched$tip.label[55] <- "Dacrydium_cupressinum" 


### checking names (the difference should equal 0 to match with studied taxa)
setdiff(my_taxa$tip_label, my_phylo_branched$tip.label)

### # Generate variance covariance matrix based on a Brownian model of evolution
# to obtain species relatedness (TOL)

phylo_matrix <- vcv(my_phylo_branched,cor=T) 

plot(my_phylo_branched)
# ape::axisPhylo()

```

META-ANALYSIS 
variance calculations

```{r calculating variance and weights}

####### use the proper sample size for all records with number of years < 4 
meta_INC$sample_size = coalesce(meta_INC$`sample_size_ifY=1`,meta_INC$number_of_years)

### calculate standard errors and weights (weight is equivalent to 1/var_z)

meta_INC$var_z <- 1/(meta_INC$sample_size - 3)
meta_INC$weight <- (meta_INC$sample_size - 3)

### remove data with too low or to good sample size

meta_INC <- meta_INC %>%
filter(var_z > 0 & var_z != "Inf")

## subset of data using only current year observations

meta_1y <- meta_INC %>%
  filter(H1 == "zero")

### mean Z score across all species

meta_Z <- meta_1y %>%
group_by(tip_label) %>%
summarize(meanZ = mean(Z,na.rm=T),
          pop = n_distinct(pop_id),
          sumY = sum(number_of_years),
          meanY = round(num(mean(number_of_years)),digits=0 )
             )

meta_Z_sec <- meta_1y %>%
filter(growth_type == "secondary growth") %>%
group_by(tip_label) %>%
summarize(meanZ = mean(Z,na.rm=T),
          pop = n_distinct(pop_id),
          sumY = sum(number_of_years),
          meanY = round(num(mean(number_of_years)),digits=0 )
             )

### histogram of Z score for entire data frame and subset of current year
hist1 <- hist(meta_INC$Z, breaks = 20, col = "grey")

hist(meta_1y$Z, breaks = 20, col = "darkgreen")

hist(meta_INC$Z, breaks = 30, col = "darkgreen") +  geom_vline(xintercept = 0, linetype="dashed")+ 
 theme_bw()
 
```
 
 
 
```{r hypothesis one - using all data and time lags}
 
## null model with all time effects (correlations for all recorded time lags - the Z effect should be nonsignificant)

model0  <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ 1 ,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_INC)

summary(model0)

### model H1 - hypothesis ONE - with time lags restricted only to growth from current or +/- year

### filtering datasets to lags that don't exceed one year

meta_H1 <- meta_INC %>%
filter(!H1 == "NA")

meta_H1 %>%  group_by(title) %>% tally() 

meta_H1 %>%  group_by(pop_id) %>% tally() 
 
### all relevant effect sizes for the analysis

# 685 effect sizes
# 78 studies
# 79 species (tip_label)
# 191 populations
# 306 populations
# 38 genera


## adjusting names to be self-explanatory 

H1names <- tibble(old = c("zero","plus1","minus1"),
  new = c( "Unlagged effect (t)","Lagged effect (t + 1)", "Lagged effect (t - 1)")
  )

meta_H1 <- meta_H1 %>%
  left_join(H1names, by = c("H1" = "old")) %>%
  mutate(H1 = coalesce(new, H1)) %>%
  dplyr::select(-new)
 

meta_H1 <- meta_H1 %>%
  mutate(H1=factor(H1)) %>% 
  mutate(H1=fct_relevel(H1,c("Lagged effect (t + 1)","Unlagged effect (t)","Lagged effect (t - 1)"))) %>%
 arrange(H1)
 
 
### rma model including conservative phylogenetic relatedness only (NULL MODEL with phylo_corMatrix)
### model H1 - the negative correlations should prevail, but are dispersed across data

model_H1_null <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ 1,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_H1)

summary(model_H1_null)

## check study heterogeneity and R2 of the model
i2_ml(model_H1_null)
r2_ml(model_H1_null) 

### rma model testing first hypothesis with time lags (H1 as a categorical variable in mods) with the conservative phylogenetic relatedness included

model_H1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ H1 - 1 ,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_H1)

summary(model_H1)
coef(summary(model_H1))
model_H1
## check study heterogeneity and R2 of the model
i2_ml(model_H1) 
r2_ml(model_H1) 

## calculate percentage of variance explained by random effects
0.0149/ (0.0149+0.0264)
0.0264/ (0.0149+0.0264)
 
model_H1$sigma2

### intercept at 0
### variance explained by study_id and effect_size_id, but not by phylogeny
### trade-off is present only in current year (highly significant) but T-1 and T+1 are not different from 0
### correlations at T-1 and T+1 are not different from 0

## a complementary model without phylogenetic relatedness (only species effect of tip_label)

model_H1_noCor <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ H1 - 1 ,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_H1)

summary(model_H1_noCor)


### intercept at 0
### variance explained by study_id and effect_size_id, but not by phylogeny
### trade-offs are present in current year and positive correlations are present in T-1 (highly significant)
### correlations at T+1 are not different from 0

### rma model including  phylogenetic relatedness only (MODEL with rotl phylo_matrix )

modelH1_rotl <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ H1 - 1,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_H1)

summary(modelH1_rotl)

## check study heterogeneity and R2 of the model
i2_ml(modelH1_rotl) 
r2_ml(modelH1_rotl) 

## calculate percentage of variance explained by random effects
0.0080/ (0.0080+0.0255)
0.0255/ (0.0080+0.0255)
 
### intercept at 0
### variance explained by study_id and effect_size_id, but not by phylogeny
### trade-offs are present in current year and positive correlations are present in T-1 (highly significant)
### correlations at T+1 are not different from 0

#extraction of marginal estimates for time lag type from the model:

res_colour_type <- orchaRd::mod_results(modelH1, mod = "H1",  group = "title") 
res_colour_type
 
### comparison of AIC of the models - all models are valid
AIC(modelH1_null,modelH1,modelH1_noCor,modelH1_corM)

```


### plot the results from the model H1 with categorical variable of time lags (rma model with phylogeny)
### orchard plot (ggplot-linked package to plot results of meta-analyses) 
 
```{r figure2}
## set my theme for figures

my_theme = theme_set(theme_minimal(base_size = 20))  # set your base text size here (e.g., 11–14)

my_theme = theme_update(
  plot.title      = element_text(face = "bold", hjust = 0, margin = margin(b = 6),size = 22),
  plot.subtitle   = element_text(margin = margin(b = 6)),
  axis.title      = element_text(),
  axis.text       = element_text(),
  legend.title    = element_text(),
  legend.text     = element_text(),
  strip.text      = element_text(face = "bold"),
  plot.margin     = margin(8, 8, 8, 8)    # extra breathing room so titles never get clipped
)

## plot the results
plot2b <- orchard_plot(model_H1 , mod = "H1", group = "title", xlab = "Effect size (Z-score)", transfm = "none", alpha = 0.3,
                       angle = 0,   fill = T, trunk.size = 0.6, twig.size = 0.4,  branch.size = 2, legend.pos = "top.out",  
                       condition.lab = NULL , flip = TRUE, g = FALSE, k=TRUE ) +
theme(legend.direction = "horizontal")   +
scale_colour_manual(values=c("darkblue","darkgreen","darkorange3")) +
scale_fill_manual(values=c("deepskyblue","chartreuse3","darkorange"))   + labs(title = "B") +   
theme(axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
      plot.title = element_text(size=25) )   
     
plot2b
 
ggsave("DATA_files/plots/H1plot.pdf", height = 9, width = 10,  bg="transparent")
    
### plot the legend

plot2b_leg <- orchard_plot(model_H1 , mod = "H1", group = "title", xlab = "Effect size (Z-score)", transfm = "none", alpha = 0.3,
                       angle = 0,   fill = T, trunk.size = 0.8, twig.size = 0.4,  branch.size = 2, legend.pos = "top.out",
                       condition.lab = NULL , flip = TRUE, g = FALSE, k=TRUE ) +theme_minimal(base_size = 24) +
theme(legend.direction = "horizontal")  
  
leg_2b <- ggpubr::get_legend(plot2b_leg)
 
ggsave("DATA_files/plots/legend_plotH1.pdf", height = 9, width = 10,  bg="transparent")
ggsave("plots/legend_plotH1.pdf", height = 9, width = 10,  bg="transparent")
    

 
## plot the results including all effect sizes with caterpillar plot (forest plot from orchard2.0)

resultsmodelH1 <- orchaRd::mod_results(model_H1, mod = "1", at = NULL,  group = "title")
 
orchaRd::caterpillars(resultsmodelH1, mod="1", xlab = "Standardised mean difference") 

## plot with results from the alternative model with rotl phylogeny

plot2b_SUP <- orchard_plot(modelH1_rotl , mod = "H1", group = "title", xlab = "Effect size (Z-score)", transfm = "none", alpha = 0.3,
                       angle = 0,   fill = T, trunk.size = 0.6, twig.size = 0.4,  branch.size = 2, legend.pos = "top.out",  
                       condition.lab = NULL , flip = TRUE, g = FALSE, k=TRUE ) +
theme(legend.direction = "horizontal")   +
scale_colour_manual(values=c("darkblue","darkgreen","darkorange3")) +
scale_fill_manual(values=c("deepskyblue","chartreuse3","darkorange"))   + # labs(title = "simplified phylogeny (tree of life)") + 
theme(plot.title = element_text(size=20), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
       )   
 
 plot2b_SUP

```

```{r distribution of correlations according to time lag}

### all data filtered to H1 

meta_INC %>%
# filter(H1 != "NA") %>%
  ggplot(aes(x=Z, color= H1, fill= H1)) +
    geom_density(alpha=0.4 ) +
    scale_fill_viridis(discrete=TRUE) +
    scale_color_viridis(discrete=TRUE)  +
   geom_vline(xintercept = 0, linetype="dashed") +
theme_bw()+ # theme_ipsum() +
    theme(
      legend.position="right") +
    ylab("Density") +
    xlab("Distribution of Z score")


### all data filtered to H1 

meta_INC %>%
filter(H1 != "NA") %>%
  ggplot(aes(x=Z, color= time_effect, fill= time_effect)) +
    geom_density(alpha=0.4 ) +
    scale_fill_viridis(discrete=TRUE) +
    scale_color_viridis(discrete=TRUE)  +
   geom_vline(xintercept = 0, linetype="dashed") +
theme_bw()+ # theme_ipsum() +
    theme(
      legend.position="right") +
    ylab("Density") +
    xlab("Distribution of Z score")


```    
     



```{r histogram of effect sizes on relevant filtered data - only current year}



### calculate average by species, study and location 

mean_res <- meta_INC %>%
  group_by(authors, species, location) %>%
  summarize(Z = (sum(weight * Z)/sum(weight)),
            var = 1/sum(weight))

## histogram of Z values across all observations
hist(mean_res$Z, breaks = 20, col = "darkgreen")

### secondary growth prevails

### mean per species per study
mean_sp_res <- meta_INC %>%
  group_by(title, tip_label) %>%
  filter(H1=="zero") %>%
  summarize(Z = (sum(weight * Z)/sum(weight)),
            var_z = 1/sum(weight))

model_sp_res <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  1,
                       random = list(~1|tip_label,
                                      ~1|title),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = mean_sp_res)

resultsmodel_sp_res <- orchaRd::mod_results(model_sp_res, mod = "1", at = NULL,  group = "title")
 
orchaRd::caterpillars(resultsmodel_sp_res, mod="1", xlab = "Standardised mean difference") 

### mean per species only

mean_species <- meta_INC %>%
 group_by(tip_label) %>%
  filter(H1=="zero") %>%
  summarize(Z = (sum(weight * Z)/sum(weight)),
            var_z = 1/sum(weight))

model_species <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  1,
                       random = list(~1|tip_label),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = mean_species)

resultsmodel_species <- orchaRd::mod_results(model_species, mod = "1", at = NULL,  group = "tip_label")
 
## draw a species-level caterpillar plot (summarised effect sizes)
orchaRd::caterpillars(resultsmodel_species, mod="1", xlab = "Standardised mean difference") 

```
 
```{r upload bioclimatic variables from worldclim }

### upload climatic data from worldclim
worldclim <- read_csv(here("DATA_files", "worldclim_trade-offs_meta.csv")) 

## check whether the coordinates match between datasets
setdiff(meta_INC$latitude, worldclim$latitude)
setdiff(meta_INC$longitude, worldclim$longitude)
 
### grouping argument
by_cols <- c("longitude", "latitude") 

# Inspect classes
sapply(meta[by_cols], class)
sapply(worldclim[by_cols], class)

# See which rows from df1 fail to find a match in df2
miss_1_in_2 <- anti_join(meta_INC, worldclim, by = by_cols)
miss_2_in_1 <- anti_join(worldclim, meta_INC, by = by_cols)

miss_1_in_2 %>% slice_head(n = 10)
miss_2_in_1 %>% slice_head(n = 10)

# normalize keys in a function to match all records from two datasets
norm_keys <- function(d, keys) {
  d %>%
    mutate(across(all_of(keys), \(x) {
      # convert types first
      if (inherits(x, "factor")) x <- as.character(x)
      # normalize unicode (NFC), swap NBSP to space, trim, squish, lower-case
      if (is.character(x)) {
        x <- stri_trans_nfc(x)
        x <- chartr("\u00A0", " ", x)
        x <- trimws(gsub("\\s+", " ", x))
        x <- tolower(x)
      }
      # round numerics used as keys (if applicable)
      if (is.numeric(x)) x <- round(x, 6)
      x
    }))
}
 
meta_INC <- norm_keys(meta_INC, by_cols)
worldclim <- norm_keys(worldclim, by_cols)
 
## control check whether the longitude is consistent across datasets
setdiff(meta_INC$longitude, worldclim$longitude)

### worldclim scaling

worldclim <-  worldclim[,-1 ] 

worldclim <- worldclim[!duplicated(worldclim), ]

exclude_cols <- c("longitude", "latitude")

# Scale all other columns
worldclim_sc <- worldclim %>%
  mutate(across(
    .cols = -all_of(exclude_cols),
    .fns = ~ scale(.)[, 1]  # scale returns a matrix, so extract the vector
  ))

str(worldclim)

### upload VPD extracted from terraclimate
vpd_locations <- read.csv(here("DATA_files","vpd-locations_trade-offs_meta.csv"))

### merging data frames (scaled values and non-scaled)
meta_FINAL_wc <- left_join(meta_INC, worldclim_sc, by =c("longitude", "latitude")) 

meta_FINAL_nonsc <- left_join(meta_INC, worldclim, by =c("longitude", "latitude")) 

meta_FINAL  <- meta_FINAL_wc  %>%  
  left_join(vpd_locations, by = join_by(replicate_site == replicate_site, latitude == lat, longitude == lon  )) 


```

 


```{r  functional traits at the species level}
### upload trait data of different species

speciesTraits <- read_csv(here("DATA_files", "speciesTraits_trade-offs_meta.csv")) %>% dplyr::select(-1) %>%
mutate(tip_label = str_replace(Species,' ','_')) %>% rename(genus = Genus) %>% 
  as.data.frame()

#add wood density of two genera:
speciesTraits <- speciesTraits %>%
  mutate(wood_density_imp = case_when(
    genus == "Cassiope" ~ paste0("0.836"),
    genus == "Fouquieria" ~ paste0("0.327") ,
  .default = as.character(wood_density_imp) ) )
 
### fill the missing data for SLA by reversing LMA

for (i in seq_along(speciesTraits$SLA)) {
  if (is.na(speciesTraits$SLA[i])) {
    speciesTraits$SLA[i] <- (1 / speciesTraits$LMA[i]) 
  }
}

# Fouquieria,diguetii,,0.326666667 wood density grams/m
# Cassiope tetragona: 0.836
 

str(speciesTraits)
speciesTraits$wood_density_imp <- as.numeric(speciesTraits$wood_density_imp)

colnames(speciesTraits)

speciesTraits_log_sc <- speciesTraits %>%  
  dplyr::select(where(is.numeric) & !c(P50)) %>%
mutate_all(~ log(.)) %>% mutate(P50 = speciesTraits$P50, 
tip_label = speciesTraits$tip_label, genus = speciesTraits$genus) %>%
  mutate(across(where(is.numeric), ~ scale(.)[, 1]))
  
## scale traits in dataframe but without log-transformation
speciesTraits_scaled <- speciesTraits %>%
  mutate(across(where(is.numeric), ~ scale(.)[, 1]))

#merging data frames

meta_nonsc_trait <-left_join(meta_FINAL_nonsc,speciesTraits, by = c("tip_label", "genus"))

meta_FINAL <- left_join(meta_FINAL,speciesTraits_log_sc, by = c("tip_label", "genus"))


#### CVp adjusted from papers and non-logged and non-scaled from mastree

meta_nonsc_trait <- meta_nonsc_trait %>% 
mutate(meanCVp = coalesce(as.numeric(CVp),mean_CVp))

## NAs introduced due to missing information on CVp
 
meta_FINAL$meanCVp = meta_nonsc_trait$meanCVp


### adding variable of the median year of the study and last year of the study
 
meta_FINAL$median_year <- sapply(strsplit(meta_FINAL$years, "-"), function(x) {
  years <- as.numeric(x)
  floor(mean(years))  # Or use round(mean(years)) if you prefer rounding
})

meta_FINAL$max_year <- sapply(strsplit(meta_FINAL$years, "-"), function(x) {
  years <- as.numeric(x)
  floor(max(years))  # Or use round(mean(years)) if you prefer rounding
})

str(meta_FINAL)

### subseting data to current year only

meta_FINAL_1y <- meta_FINAL %>%
filter(H1 == "zero")

meta_nonsc_trait_1y <- meta_nonsc_trait %>%
filter(H1 == "zero")

```


### rma model testing first hypothesis with time lags (H1 as a categorical variable in mods) with the conservative phylogenetic relatedness included

### intercept at 0
### variance explained by study_id and effect_size_id, but not by phylogeny
### trade-off is present only in current year (highly significant) but T-1 and T+1 are not different from 0
### correlations at T-1 and T+1 are not different from 0

## a complementary model without phylogenetic relatedness (only species effect of tip_label)

all tested models follow the same structure:
 
model0.1 = the main model including phylogenetical relatedness based on the Zanne et al.2016 paper with conservative divergence time
model0.2 = a coplementary model with species random effect instead of phylogenetic matrix
model0.3 = an extra model including phylogenetical relatedness from open tree of life (rotl) with relative divergence time (Grafen's method)
  
 
```{r FULL and FINAL MODELS with all variables }

###        CURRENT YEAR OBSERVATIONS ONLY

### rma model testing further hypotheses excluding time lags to keep the main biological meaning 
## model with conservative phylogenetic relatedness
 

modelFINALnull <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ 1 ,
                    random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                         control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelFINALnull)

## check study heterogeneity and R2 of the model
i2_ml(modelFINALnull) 
r2_ml(modelFINALnull) 

orchard_plot(modelFINALnull, group = "title",  xlab = "Z scores")

##  mean Z value is significantly lower than 0 (trade-offs detected)
### large amount of variance explained by phylogeny and study_id


## null rma model with rotl phylogeny
## large amount of variance explained by phylogeny = trade-off non-detected despite negative estimate

###################

#### rma model without phylogeny 

NULLmodel0.1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ 1,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)

summary(NULLmodel0.1)


orchard_plot(NULLmodel0.1, group = "title",  xlab = "Z scores")


## model with another phylogeny 

NULLmodel0.2 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ 1,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)

summary(NULLmodel0.2)

orchard_plot(NULLmodel0.2, group = "title",  xlab = "Z scores")

##  mean Z value is non-significantly lower than 0 (trade-offs undetected)
### large amount of variance explained by phylogeny and study_id


### compare AIC of null models

AIC(modelFINALnull,NULLmodel0.1,NULLmodel0.2)

### models with phylogeny have slightly better AIC value

## old codes
# sp_eff <- blup(NULLmodel0.1)$tip_label   # pseudo-code; extract species random effects/BLUPs
# phytools::phylosig(phy, sp_eff, method = "lambda", test = TRUE)

### figures - caterpillar plots

model_resultsNULL <- orchaRd::mod_results(modelFINALnull, mod = "1", at = NULL,  group = "title")
 
  orchaRd::caterpillars(model_resultsNULL, mod="1", xlab = "Standardised mean difference")  + 
theme(axis.text.x =element_text(size = 16), 
           axis.title=element_text(size = 16),
           legend.text = element_text(size=14))

## model with no phylogeny

model_results0.1 <- orchaRd::mod_results(NULLmodel0.1, mod = "1", at = NULL,  group = "title")
 
orchaRd::caterpillars(model_results0.1, mod="1", xlab = "Standardised mean difference") 
  
## model with phylogeny matrix  

model_results0.2 <- orchaRd::mod_results(NULLmodel0.2, mod = "1", at = NULL,  group = "title")
 
orchaRd::caterpillars(model_results0.2, mod="1", xlab = "Standardised mean difference")
    

 ### plot with current year effect size (null final model) 

plot1b <- orchaRd::caterpillars(model_resultsNULL, mod="1", xlab = "Effect size (Z-score)",  g = FALSE, k = FALSE, overall = FALSE,  colerrorbar = "darkgreen" , colpoly = "darkorange" , colpoint = "#ffe699")   + labs(title="B")  +
theme(axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 18),
           legend.text = element_text(size=20),
          axis.text.y =element_text(size = 20),
         plot.title  = element_text(size=24))

#FIGURE1b
plot1b
 
### old plot showing lagged effects and pulled current year effect
print(plot2b + plot1b)


 
## 2) Arrange side-by-side with shared legend (optional)
# guides = "collect" pulls legends together; adjust widths so neither title is squashed

plot1 <- (plot1a + plot1b ) +
  plot_layout(widths = c(1, 1)) +
my_theme


plot1

## 3) Save to a high-quality PDF (vector, embedded fonts)
# Pick a sensible width/height; ~16x7 inches works well for two side-by-side panels.
# Use device = cairo_pdf for robust font embedding across platforms.

ggsave(
  filename = "plot1.pdf",
  plot     = plot1,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 16,
  height   = 9,
  units    = "in"
)


combined <- (plot1a + plot1b ) +
  plot_layout(guides = "collect", widths = c(1, 1)) +
my_theme

#ggsave("H1_plot.pdf", height = 12, width = 12)
#dev.off()

plot1bSOLO <- orchaRd::caterpillars(model_results0.1, mod="1", xlab = "Effect size (Z-score)", g = FALSE, overall = FALSE,  colerrorbar = "darkgreen" , colpoly = "black" , colpoint = "#ffe699")   + labs(title="B") +
theme(axis.text.x =element_text(size = 30), 
           axis.title=element_text(size = 26),
           legend.text = element_text(size=30),
          axis.text.y =element_text(size = 30),
         plot.title  = element_text(size=24))

plot1bSOLO
 

ggsave(
  filename = "plot1bSOLO.png",
  plot     = plot1bSOLO,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 12,
  height   = 10,
  units    = "in"
)



```


```{r new data frame for testing lags}

###  grouping data according to the population and lag

meta_lags_long <- meta_INC %>%
filter(H1 != "NA") %>% 
    group_by(pop_id,tip_label,replicate_site,title,H1) %>%
reframe(pop_id = unique(pop_id),
meanZ = mean(Z,na.rm=T),
)
 
#### 

meta_lags_final <- dcast(
  meta_lags_long,
 title +  replicate_site + tip_label + pop_id   ~ H1,   # rows = site × species, columns = category
  value.var = "meanZ"         ,  # what to put into the cells,
)

 

### adding value from minus one and plus one year lag to the dataset

meta_FINAL_1y <- meta_FINAL_1y %>%
  left_join(meta_lags_final, by = join_by(pop_id == pop_id , title == title, tip_label == tip_label, replicate_site == replicate_site)) 

## zero is Z value
## minus 1 is the correlation between growth and reproduction from the previous year
## plus1 is the correlation between growth and reproduction in the following year


```


 
```{r  within species correlations between effect sizes according to the lags}
 
###### testing lag effects within observations 
###### are trade-offs stronger if correlations in T-1 and T+1 are positive/negative
 
inter_lags_model <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  minus1 + plus1,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(inter_lags_model)    
 
  
## check study heterogeneity and R2 of the model
i2_ml(inter_lags_model) 
r2_ml(inter_lags_model) 

### the more positive correlations in T-1 strengthen the trade-offs between reproduction and growth (= more negative correlations in T)
## much variance explained by study_id

## model0.1

inter_lags_model0.1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  minus1 + plus1,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(inter_lags_model0.1)    

###
                
inter_lags_model0.2 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  minus1 + plus1,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(inter_lags_model0.2)   

## check study heterogeneity and R2 of the model
i2_ml(inter_lags_model0.2) 
r2_ml(inter_lags_model0.2) 

### change in effect sizes according to the plant group

inter_lags_model1.0 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                          mods = ~   minus1 ,
                      # mods = ~  minus1 * plant_type, 
                       # mods = ~  minus1 * meanCVp, 
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(inter_lags_model1.0)    

### exploratory plots
  
## scatter plot of values between two years
plot(meta_FINAL_1y$minus1,meta_FINAL_1y$plus1)

bubble_interlag_types <- orchaRd::mod_results(inter_lags_model0.1, mod = "minus1", group = "title",
                    weights = "prop")


orchaRd::bubble_plot(bubble_interlag_types, group = "title",  mod = "minus1", xlab = "minus1", legend.pos = "top.left") + 
geom_point(data = meta_FINAL_1y, aes(x =  minus1, y = Z, 
                                     color = plant_type, size = 1/sqrt(var_z)), alpha = 0.6) +  scale_colour_manual(values=c("darkorange","black")) 

##
## plot with autocorrelations and different CVp values
 
bubble_interlag_cvp <- orchaRd::mod_results(inter_lags_model0.1, mod = "minus1", group = "title",
                    weights = "prop")
 

orchaRd::bubble_plot(bubble_interlag_cvp, group = "title",  mod = "minus1", xlab = "minus1", legend.pos = "top.left") + 
geom_point(data = meta_FINAL_1y, aes(x =  minus1, y = Z,  color = meanCVp, size = 1/sqrt(var_z)), alpha = 0.6)   + scale_colour_viridis_b()




## FIGURE 2
## final plot with significant effect of Z from previous year
 

bubble_interlag <- orchaRd::mod_results(inter_lags_model, mod = "minus1", group = "title",
                    weights = "prop")


orchaRd::bubble_plot(bubble_interlag, group = "title",  mod = "minus1", xlab = "minus1", legend.pos = "top.left", est.col = "grey",  
  ci.col = "red",
  pi.col = "blue",) + 
geom_point(data = meta_FINAL_1y, aes(x = minus1, y = Z,   size = 1/sqrt(var_z)), alpha = 0.2)   
  

plot2leg <- orchaRd::bubble_plot(bubble_interlag, group = "title",  mod = "minus1", xlab = "minus1", legend.pos = "top.right") +  scale_fill_manual(values=c("darkred")) +   theme(legend.direction = "vertical")   + theme(axis.text.y =element_text(size = 14), 
       axis.text.x =element_text(size = 16), 
           axis.title=element_text(size = 16),
           legend.text = element_text(size=14),
           legend.title = element_text(size=12)
          ) + labs(x="Z-score in t-1", y = "Z-score in t", title = "C") +
  theme(aspect.ratio = (1))

plot2leg

leg <- ggpubr::get_legend(plot2leg)


## FIGURE 2C

plot2c <- orchaRd::bubble_plot(bubble_interlag, group = "title",  mod = "minus1", xlab = "minus1", legend.pos = "none", k.pos = "bottom.left") +  scale_fill_manual(values=c("darkred")) +   theme(legend.direction = "vertical")   + labs(x="Z-score in t-1", y = "Z-score in t", title = "C") + theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))

plot2c
  
  geom_point(data = meta_FINAL_1y, aes(x = minus1, y = Z,  size = 1/sqrt(var_z)), alpha = 0.5) +  theme(legend.direction = "vertical")   +
#scale_colour_manual(values=c( "darkgreen")) +
scale_fill_manual(values=c("orange"))   + 
theme(axis.text.y =element_text(size = 14), 
       axis.text.x =element_text(size = 16), 
           axis.title=element_text(size = 16),
           legend.text = element_text(size=14),
           legend.title = element_text(size=12)
          ) + 
  theme(aspect.ratio = (1))
     

## final plot with nonsignificant effect of Z between T0 and T1 (subsequent year)
#### FIGURE 2D

bubble_interlag2 <- orchaRd::mod_results(inter_lags_model, mod = "plus1", group = "title",
                    weights = "prop")


plot2d <- orchaRd::bubble_plot(bubble_interlag2, group = "title",  mod = "plus1", xlab = "plus1", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") +  scale_fill_manual(values=c("grey")) + labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))

plot2d

plot2 <- plot2b + (plot2c/plot2d)  +
  plot_layout(widths = c(1, 1)) +
my_theme

plot2

ggsave(
  filename = "plot2.pdf",
  plot     = plot2,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 16,
  height   = 10,
  units    = "in"
)

ggsave(
  filename = "plot2.png",
  plot     = plot2,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 16,
  height   = 10,
  units    = "in"
)

## check other size 

ggsave(
  filename = "plot2.pdf",
  plot     = plot2,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 15,
  height   = 10,
  units    = "in"
)
plot2b_SOLO <- plot2b + theme(axis.text.y =element_text(size = 25), 
       axis.text.x =element_text(size = 25), 
           axis.title=element_text(size = 25),
           legend.text = element_text(size=25),
      plot.title = element_text(size=25) )   
     
plot2b_SOLO  
 
plot2b_SUP

ggsave(
  filename = "plot2b_SUP.pdf",
  plot     = plot2b_SUP,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 12,
  height   = 10,
  units    = "in"
)


ggsave(
  filename = "plot1aSOLO.png",
  plot     = plot1aSOLO,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 10,
  height   = 10,
  units    = "in"
)


plot2SOLO <- (plot2c + plot2d)  + theme(axis.text.y =element_text(size = 25), 
       axis.text.x =element_text(size = 25), 
           axis.title=element_text(size = 25),
           legend.text = element_text(size=25),
      plot.title = element_text(size=25) )   


```

####### none of the tested extra models showed significant effect of climate and traits (PCA of all traits at the end of the script)

```{r correlations between traits }

cor.test(meta_FINAL_1y$seed_mass, meta_FINAL_1y$Height) 
cor.test(meta_FINAL_1y$seed_mass, meta_FINAL_1y$wood_density_imp) 
cor.test(meta_FINAL_1y$seed_mass, meta_FINAL_1y$SLA) 
cor.test(meta_FINAL_1y$seed_mass, meta_FINAL_1y$`Annual Mean Temperature`)

cor.test(meta_FINAL_1y$wood_density_imp, meta_FINAL_1y$Height) 
cor.test(meta_FINAL_1y$SLA, meta_FINAL_1y$Height)
cor.test(meta_FINAL_1y$`Annual Mean Temperature`,  meta_FINAL_1y$Height)

cor.test(meta_FINAL_1y$SLA, meta_FINAL_1y$Height) 
cor.test(meta_FINAL_1y$wood_density_imp, meta_FINAL_1y$SLA) 

cor.test(meta_FINAL_1y$meanCVp, meta_FINAL_1y$SLA) 
cor.test(meta_FINAL_1y$meanCVp, meta_FINAL_1y$wood_density_imp) 

## no correlation between CVp and traits
cor.test(meta_FINAL_1y$seed_mass, meta_FINAL_1y$meanCVp) 
cor.test(meta_FINAL_1y$meanCVp, meta_FINAL_1y$Height) 

### all species - raw data
cor.test(speciesTraits$seed_mass, speciesTraits$Height) 

str(meta_FINAL_1y)
### CORRELATION BETWEEN CLIMATIC VARIABLES

cor.test(meta_FINAL_1y$`Annual Mean Temperature`,  meta_FINAL_1y$`Max Temperature of Warmest Month`)
cor.test(meta_FINAL_1y$`Annual Mean Temperature`,  meta_FINAL_1y$`Min Temperature of Coldest Month`)

cor.test(meta_FINAL_1y$`Annual Mean Temperature`,  meta_FINAL_1y$`Mean Temperature of Warmest Quarter`)
cor.test(meta_FINAL_1y$`Annual Mean Temperature`,  meta_FINAL_1y$`Mean Temperature of Coldest Quarter`)

```

```{r  methods etc}

## non-significant effects of methods, biomes and plant type (a few models that do not belong to the manuscript)
### model 1.0 - around time lags and different methods

modelPLUS <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ H1 - 1 + plant_type + growth_type + reproduction_type + scale + biome ,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL)


coef(summary(modelPLUS))

orchard_plot(modelPLUS, mod = "H1", group = "title", by="scale", xlab = "Correlation coefficient", transfm = "none", alpha = 0.4, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = "Method") + theme(legend.direction = "vertical")  

### 

modelPLUS1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  plant_type + growth_type + reproduction_type + scale + biome ,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)


summary(modelPLUS1)

###
### methods only

modelPLUS2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    growth_type + reproduction_type + scale + plant_type,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_INC)
                        
summary(modelPLUS2)

#######

orchard_plot(modelPLUS2, mod = "growth_type", group = "title", by="scale", xlab = "Correlation coefficient", transfm = "none", alpha = 0.4, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = "Method") + theme(legend.direction = "vertical")  

orchard_plot(modelPLUS2, mod = "reproduction_type", group = "title", by="scale", xlab = "Correlation coefficient", transfm = "none", alpha = 0.4, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = "Method") + theme(legend.direction = "vertical")  

###

orchard_plot(modelPLUS1, mod = "scale", group = "title", by="plant_type", xlab = "Correlation coefficient", transfm = "none", alpha = 0.4, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = "Method") + theme(legend.direction = "vertical")  

###


modelH1_PLUS1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ H1 -1 + plant_type + scale +  growth_type ,
                    random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                    control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL)

summary(modelH1_PLUS1)

### orchard plot with plant types

orchard_plot(modelH1_PLUS1, mod = "growth_type", group = "title", by="growth_type", xlab = "Correlation coefficient", transfm = "none", alpha = 0.4, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = "Method") + theme(legend.direction = "vertical")  
 

##  MODEL testing different scales and methods (in the supplement)

model_methods <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    growth_type + scale  + reproduction_type + plant_type,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(model_methods)

## check study heterogeneity and R2 of the model
i2_ml(model_methods) 
r2_ml(model_methods) 

## calculate percentage of variance explained by random effects
0.0558 / (0.0558+0.0463)
0.0463 / (0.0558+0.0463)

####### sensitivity analysis (models 0.1 without phylogeny)
model_methods0.1 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    growth_type + scale  + reproduction_type + plant_type,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)

## model 0.2 based on the syntethic tree
model_methods0.2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    growth_type + scale  + reproduction_type + plant_type,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)

####### 
summary(model_methods0.1)     
summary(model_methods0.2)

## check study heterogeneity and R2 of the model
i2_ml(model_methods0.2) 
r2_ml(model_methods0.2) 

## calculate percentage of variance explained by random effects
0.0874 / (0.0874 + 0.0468)

####### consistent patterns across all models

plotM1 <- orchard_plot(model_methods, mod = "growth_type", group = "title",  xlab = "Z-score", transfm = "none", alpha = 0.5, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = NULL, k = TRUE, g = FALSE) + theme(legend.direction = "horizontal") + labs(title = "A") + 
  scale_fill_manual(values=c("purple3","darkgreen","brown4")) + 
  scale_color_manual(values=c("purple4","green3","brown2")) +
  theme(axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 18),
           legend.text = element_text(size=20),
          axis.text.y =element_text(size = 20),
         plot.title  = element_text(size=24))

plotM2 <-  orchard_plot(model_methods, mod = "reproduction_type", group = "title",  xlab = "Z-score", transfm = "none", alpha = 0.5, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = NULL,  g = FALSE) + theme(legend.direction = "horizontal") + labs(title = "B") + scale_fill_manual(values=c("darkgreen","goldenrod2","red")) + scale_color_manual(values=c("green3","goldenrod1","red4")) +
  theme(axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 18),
           legend.text = element_text(size=20),
          axis.text.y =element_text(size = 20),
         plot.title  = element_text(size=24))

plotM3 <-  orchard_plot(model_methods, mod = "scale", group = "title",  xlab = "Z-score", transfm = "none", alpha = 0.5, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = NULL,  g = FALSE) + theme(legend.direction = "horizontal")  +    labs(title = "C") + scale_fill_manual(values=c("blue2","green3")) + scale_color_manual(values=c("blue4","darkgreen")) +
  theme(axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 18),
           legend.text = element_text(size=20),
          axis.text.y =element_text(size = 20),
         plot.title  = element_text(size=24))

plotM4 <-  orchard_plot(model_methods, mod = "plant_type", group = "title", xlab = "Z-score", transfm = "none", alpha = 0.5, angle = 0,  cb = FALSE, legend.pos = "top.left", condition.lab = NULL,  g = FALSE) + theme(legend.direction = "horizontal") + labs(title = "D") + scale_fill_manual(values=c("darkorange","black")) + scale_color_manual(values=c("darkorange","black")) +
  theme(axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 18),
           legend.text = element_text(size=20),
          axis.text.y =element_text(size = 20),
         plot.title  = element_text(size=24))

## plot FIGURE S6
methods_plot <- (plotM1 + plotM2 ) / (plotM3 + plotM4) 

methods_plot

# save FIGS6
ggsave(
  filename = "methods_orchard-plot.pdf",
  plot     =  methods_plot,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 16,
  height   = 12,
  units    = "in"
)

ggsave(
  filename = "methods_orchard.png",
  plot     =  methods_plot,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 16,
  height   = 12,
  units    = "in"
)


```

 
 
```{r  FINAL MODELS AND THEIR COEFFICIENT PLOTS} 

### FINAL MODEL AND FIGURE 4

##### climate interaction statistics in the model below

FINALmodel_H3_int  <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   `Annual Mean Temperature` * `Annual Precipitation` + vpd +
                        SLA + wood_density_imp  + plant_height + seed_mass,   
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(FINALmodel_H3_int) 

## nonsignificant interaction ---> best model with additive effects only

FINALmodel_H3  <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   `Annual Mean Temperature` + `Annual Precipitation` + vpd +
                        SLA + wood_density_imp  + plant_height + seed_mass,   
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                         R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
 summary(FINALmodel_H3) 

## check study heterogeneity and R2 of the model
i2_ml(FINALmodel_H3) 
r2_ml(FINALmodel_H3) 

## calculate percentage of variance explained by random effects
0.0418 / (0.0418 + 0.0543) # phylogeny explains 43% of random effect variation

## nonsignificant effects in the full model

## model reduced to additive effects without phylogeny 

FINALmodel_H3_0.1 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    `Annual Mean Temperature` + `Annual Precipitation` + vpd + SLA + wood_density_imp  + plant_height + seed_mass,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id), 
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(FINALmodel_H3_0.1)       

## non-significant effects in the full simplified model

## model reduced to additive effects with phylogeny from open tree of life

FINALmodel_H3_0.2 <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    `Annual Mean Temperature` + `Annual Precipitation` + vpd + SLA + wood_density_imp  + plant_height + seed_mass,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id), 
                         R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(FINALmodel_H3_0.2)         

## check study heterogeneity and R2 of the model
i2_ml(FINALmodel_H3_0.2) 
r2_ml(FINALmodel_H3_0.2) 

## calculate percentage of variance explained by random effects
0.0641 / (0.0641 + 0.0535) #  phylogeny explains 55% of random effect variation
       
## non-significant effects in the full model with rotl phylogeny

## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## 
## model reduced to additive effects only and with CVp (but on reduced number of 48 species compared to the full model with 79)

modelCV_H3_1.0  <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  meanCVp + `Annual Mean Temperature` + `Annual Precipitation` + 
                          SLA + wood_density_imp  + plant_height + seed_mass ,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(modelCV_H3_1.0) 

## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## ## 
## FIGURE 4
### coefficient plot for model H3

coefFINALmodel <- coef(summary(FINALmodel_H3))

rownames(coefFINALmodel) <-  c ( "Intercept" , "Mean Annual Temperature" ,"Mean Annual Precipitation" , "Mean VPD", "SLA", "Wood density",  "Plant height"  ,  "Seed mass"  )

coefFINALmodel <- coefFINALmodel %>% mutate(across(c(1:6), as.numeric) )  
 
 
c <- coefFINALmodel %>%
  ggplot(aes(y = fct_rev(rownames(coefFINALmodel)))) + 
  theme_classic()

plot_H3   <- c +  geom_point(aes(x=estimate), shape=15, size=3) +
  geom_linerange(aes(xmin=ci.lb, xmax=ci.ub))   +  geom_vline(xintercept = 0, linetype="dashed", colour ="darkgrey") +
  theme(text = element_text(size = 15)) + 
  labs(x="Z-score (± 95% CI)", y="")  + coord_cartesian(  xlim=c(-.8, .8))   +
   scale_x_continuous(breaks = c(-0.5,-0.25,0,0.25,0.5))
                            
plot_H3
 
ggsave(
  filename = "H3_plot.pdf",
  plot     =  plot_H3,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 7,
  height   = 6,
  units    = "in"
)


```

```{r TRAIT model and the effects between Z and each trait (raw data and bubble plots for scaled and non-scaled models)}

## two relevant final models

summary(FINALmodel_H3) 

# # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # #
# # # # plot based on final model with annual temperature

bubble_temp  <- orchaRd::mod_results(FINALmodel_H3, mod = "Annual Mean Temperature", group = "title",
                    weights = "prop")
                                 
H3_temp <- orchaRd::bubble_plot(bubble_temp, group = "title",  mod = " `Annual Mean Temperature`", xlab = "MAT", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x =  `Annual Mean Temperature`, y = Z,   size = 1/sqrt(var_z)), alpha = 0.2, colour="black")  +
#  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))  

# # # # plot based on final model with precipitation

bubble_prec <- orchaRd::mod_results(FINALmodel_H3, mod = "Annual Precipitation", group = "title",
                    weights = "prop")
                                 
H3_prec <- orchaRd::bubble_plot(bubble_prec, group = "title",  mod = "Annual Precipitation", xlab = "MAP", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x =  `Annual Precipitation`, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2)  + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))

 
# # # # plot based on final model with vpd

bubble_vpd <- orchaRd::mod_results(FINALmodel_H3, mod = "vpd", group = "title",
                    weights = "prop")
                                 
H3_vpd <- orchaRd::bubble_plot(bubble_vpd, group = "title",  mod = "vpd", xlab = "Mean VPD", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x =  vpd, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2)  + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))


# # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # #
# # # #  plot based on final model - SLA
bubble_SLA  <- orchaRd::mod_results(FINALmodel_H3, mod = "SLA", group = "title",
                    weights = "prop")
                                 
H3_SLA <- orchaRd::bubble_plot(bubble_SLA, group = "title",  mod = "SLA", xlab = "SLA", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x = SLA, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2)  + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))


# # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # #
# # # #  plot based on final model - WOOD DENSITY
bubble_wood <- orchaRd::mod_results(FINALmodel_H3, mod = "wood_density_imp", group = "title",
                    weights = "prop")
                                 
H3_wd <-orchaRd::bubble_plot(bubble_wood, group = "title",  mod = "wood_density_imp", xlab = "Wood density", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x = wood_density_imp, y = Z, size = 1/sqrt(var_z)), alpha = 0.2) + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))


# # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # #
# # # #  plot based on final model - HEIGHT
bubble_height  <- orchaRd::mod_results(FINALmodel_H3, mod = "plant_height", group = "title",
                    weights = "prop")
                                 
H3_height <- orchaRd::bubble_plot(bubble_height, group = "title",  mod = "plant_height", xlab = "Plant height", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x = plant_height, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2)  + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))

  # # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # #
# # # #  plot based on final model - SEED MASS

bubble_seed <- orchaRd::mod_results(FINALmodel_H3, mod = "seed_mass", group = "title",
                    weights = "prop")
                                 
H3_seed <- orchaRd::bubble_plot(bubble_seed, group = "title",  mod = "seed_mass", xlab = "Seed mass", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") + 
geom_point(data = meta_FINAL_1y, aes(x = seed_mass, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2) + 
  #  labs(x="Z-score in t+1", y = "Z-score in t", title = "D") +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))

 
# # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # ## # # # 
 ## # # #  arrange plots for FIGURE S11

raw_data_plots <-( H3_temp + H3_prec + H3_vpd  ) / (H3_SLA + H3_wd ) / ( H3_height + H3_seed)
    
raw_data_plots
 
ggsave(
  filename = "raw_data_plots.pdf",
  plot     =  raw_data_plots,
  #device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 15,
  height   = 14,
  units    = "in"
)

```



```{r year of study and cvp - hypothesis 4}

### model with CVp and year of study (change of trade-offs over time)

modelCV  <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   (meanCVp)  * (max_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV)

## check study heterogeneity and R2 of the model
i2_ml(modelCV) 
r2_ml(modelCV)

## calculate percentage of variance explained by random effects
0.0382 / (0.0382+0.0285) # phylogeny explains 57% of random variation
0.0285 / (0.0382+0.0285)

### raw data with CVp and time

bubble_cvp <- orchaRd::mod_results(modelCV, mod = "meanCVp", group = "title",
                    weights = "prop")
                                 
CVp_raw_data <- orchaRd::bubble_plot(bubble_cvp , group = "title",  mod = "meanCVp", xlab = "Mean CVp", legend.pos = "none", est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left") +
geom_point(data = meta_FINAL_1y, 
aes(x = meanCVp, y = Z, size = 1/sqrt(var_z)), alpha = 0.2)  +
theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))
 

bubble_cvp_time <- orchaRd::mod_results(modelCV, mod = "max_year", group = "title",
                    weights = "prop")
                                 
time_raw_data <- orchaRd::bubble_plot(bubble_cvp_time , group = "title",  mod = "max_year", xlab = "Final year of the study", legend.pos = "none",est.lwd = 0.5, est.col = NA, ci.col = NA ,pi.col = NA, k.pos = "bottom.left")+
geom_point(data = meta_FINAL_1y, 
aes(x = max_year, y = Z,  size = 1/sqrt(var_z)), alpha = 0.2) +
  theme(plot.title =element_text(size = 25), axis.text.y =element_text(size = 20), 
       axis.text.x =element_text(size = 20), 
           axis.title=element_text(size = 20),
           legend.text = element_text(size=20),
           legend.title = element_text(size=20) ) +
  theme(aspect.ratio = (1))


## plots
CVp_raw_data + time_raw_data

#######################################################################
## model without phylogeny
modelCV_0.1 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    (meanCVp)  * (max_year),
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV_0.1)

## model based on the syntethic tree
modelCV_0.2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   (meanCVp)  * (max_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV_0.2)

## check study heterogeneity and R2 of the model
i2_ml(modelCV_0.2) 
r2_ml(modelCV_0.2)

## calculate percentage of variance explained by random effects
0.0860/ (0.0860 + 0.0200) #  phylogeny 81
0.0200/ (0.0860 + 0.0200) # 


########### medians of year study

modelmedCV_0.2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   scale(meanCVp)  *  scale(median_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelmedCV_0.2)


modelCV_1.2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~    scale(meanCVp)  *  scale(median_year),
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV_1.2)



modelCV_2.2 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   scale(meanCVp)  * scale(median_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV_2.2)

vif(modelCV_0.2)

### plot with CVp

bubble_CV1  <- orchaRd::mod_results(modelCV, mod = "meanCVp", group = "title",
                    weights = "prop")
                    
                    
orchaRd::bubble_plot(bubble_CV1, group ="title",  mod = "meanCVp" , xlab = "CVp",  legend.pos = "top.left")   +
geom_point(data = meta_FINAL_1y, aes(x = meanCVp, y = Z, size = 1/sqrt(var_z)), alpha = 0.6) + scale_color_discrete()
  
  
  
### plot with max year
bubble_CV2  <- orchaRd::mod_results(modelCV_0.1, mod = "max_year", group = "title",
                    weights = "prop")
                    
                    
orchaRd::bubble_plot(bubble_CV2, group ="title",  mod = "max_year" , xlab = "max_year",  legend.pos = "top.left")   +
geom_point(data = meta_FINAL_1y, aes(x = max_year, y = Z, size = 1/sqrt(var_z)), alpha = 0.6) + scale_color_discrete()



## ## ## model with median year (consistent pattern) but using entire data set with H1 interaction

modelCV_0.3 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                      mods = ~ H1 *  scale(meanCVp)  *  scale(median_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                         R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL %>% filter(!H1 == "NA"))
                        
summary(modelCV_0.3)

modelCV_1.3 <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                      mods = ~   scale(meanCVp)  *  scale(median_year),
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL)
                        
summary(modelCV_1.3)

vif(modelCV_0.3)

### extra model with climatic variables

modelCV_extra <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  scale(meanCVp)  * scale(max_year) +  `Annual Mean Temperature` +  `Annual Precipitation` ,
                   random = list(~1|tip_label,
                                 ~1|title,
                                 ~1|ES_id),
                         R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
                        
summary(modelCV_extra)

### raw data across years
ggplot(data=meta_FINAL_1y, aes(x=max_year,y = Z, col=meanCVp))  +
geom_point() + geom_smooth(method = "lm", se = T)
 
```
 

```{r plot with predicted data}

summary(modelCV)

## small changes to the way the predictions are made - more dense sampling of values for year and CV

time_vals <- seq(min(meta_FINAL_1y$max_year, na.rm = TRUE),
                      max(meta_FINAL_1y$max_year, na.rm = TRUE), by = 0.1)

cvp_vals <- seq(min(meta_FINAL_1y$meanCVp, na.rm = TRUE),
                max(meta_FINAL_1y$meanCVp, na.rm = TRUE),
                by = 0.001)


newdat <- expand.grid(
  meanCVp  =  as.numeric(cvp_vals),
  max_year = as.numeric(time_vals),
  KEEP.OUT.ATTRS = FALSE
) 

#####

matriX <- model.matrix(~ meanCVp * max_year, data = newdat)
colnames(matriX)[colnames(matriX) == "(Intercept)"] <- "intrcpt"              

coef_namesX <- names(coef(modelCV))    # e.g. "intrcpt","cvp","time","cvp:time"
beta_namesX <- setdiff(coef_names, "intrcpt")
matriX <- matriX[, beta_namesX, drop = FALSE]                

pred_matrix <- predict(modelCV, newmods = matriX)         

plotdat <- cbind(newdat,
                 pred  = as.numeric(pred_matrix$pred),
                 ci.lb = as.numeric(pred_matrix$ci.lb),
                 ci.ub = as.numeric(pred_matrix$ci.ub))

df <- meta_FINAL_1y %>% ungroup()  %>% dplyr::select(max_year,meanCVp) ## give x and y for the graph

df <- na.omit(df)

convex_hull <- df %>% as_tibble() %>% dplyr::slice(chull(max_year,meanCVp)) ## give x and y for the graph

coords <- as.data.frame(convex_hull)
sp_poly <- SpatialPolygons(list(Polygons(list(Polygon(coords)), ID=1)))

mm1 <- plotdat
mm1 <- data.frame(mm1)

sp_poly_df <- SpatialPolygonsDataFrame(sp_poly, data=data.frame(ID=1))

spdf <- sp::SpatialPointsDataFrame(coords = mm1[, c("max_year","meanCVp")], data = mm1) 

whatever <- spdf[!is.na(over(spdf, as(sp_poly_df, "SpatialPolygons"))), ]
whatever <- as.data.frame(whatever)

## FINAL PLOT - FIGURE 5 in manuscript
   
    plot_CVp <-  ggplot() +
  geom_tile(data = whatever, aes(x = max_year, y = meanCVp, fill = pred)) +
  scale_fill_gradient2(name = "Z", limits = c(-1.5, 0.5), high =  "#2166AC", mid ="#F7F7F7",
low = "#B2182B", space = "Lab", midpoint = 0) +
  geom_point(data = meta_FINAL_1y, aes(x = max_year, y = meanCVp), size = 5,
             shape = 21,  colour = "black" , fill = "black", alpha = 0.4) +
  theme_classic() +
  theme(axis.text = element_text(size = 18, colour = "black"),
        axis.title.x = element_text(size = 20, colour = "black"),
        axis.title.y = element_text(size = 20, colour = "black"),
        plot.title = element_text(size = 24)) +
  theme(legend.title=element_text(size = 20), 
        legend.text=element_text(size = 18)) + 
      labs(x="Final year of the study", y = "Mean CVp", title = "A")   + theme(aspect.ratio = (1))   

 plot_CVp
 
 ggsave(
  filename = "plot4.png",
  plot     = plot_CVp,
  dpi = 800
)
 
 ggsave(
  filename = "plot4.pdf",
  plot     = plot_CVp,
  width    = 12,
  height   = 12,
  units    = "in"
)
 
```
 
 
extra plots with raw data in the supplement

```{r plot with predicted data}

###### plot with interaction shown as two linear predictions
###### interaction terms as slopes
######

## choose values density for predictions
 
  newdat_lines <- expand.grid(
  meanCVp  = as.numeric(c(0.6,1.6)),
  max_year = time_vals,
  KEEP.OUT.ATTRS = FALSE
)           
                 
 ## create prediction matrix 
  
X <- model.matrix(~ meanCVp * max_year, data = newdat_lines)
colnames(X)[colnames(X) == "(Intercept)"] <- "intrcpt"              
                 
coef_names <- names(coef(modelCV))            # e.g. "intrcpt","cvp","time","cvp:time"
beta_names <- setdiff(coef_names, "intrcpt")
X <- X[, beta_names, drop = FALSE]                
               
 pr <- predict(modelCV, newmods = X)           
               
## prepare a data frame
pred_cvp_plot <- cbind(newdat_lines,
                 pred  = as.numeric(pr$pred),
                 ci.lb = as.numeric(pr$ci.lb),
                 ci.ub = as.numeric(pr$ci.ub))


## 6) Plot two lines (one per CVp) across time with CI ribbons

pred_cvp_plot$cvp_lab <- factor(round(pred_cvp_plot$meanCVp, 2))

ggplot(pred_cvp_plot, aes(x = max_year, y = pred, color = cvp_lab, fill = cvp_lab)) +
  geom_ribbon(aes(ymin = ci.lb, ymax = ci.ub), alpha = 0.15) +
  geom_line(linewidth = 1) +  
  scale_colour_manual(values=c("darkred","darkblue")) +  
  labs(x = "Time", y = "Predicted effect size", color = "CVP", fill = "CVP") +
  theme_minimal()    
  
 
  ggplot() +
  geom_point(data = meta_FINAL_1y, aes(x = max_year, y = Z,  size = 1/sqrt(var_z), col=meanCVp), alpha = 0.8)  +
scale_color_viridis()

## final plot with raw data in the supplement (FIG.S12)
  
sup_fig12 <- ggplot() + 
  geom_line(data=pred_cvp_plot, aes(x = max_year, y = pred,   col = cvp_lab),linewidth = 1) +  
    geom_ribbon(data=pred_cvp_plot, aes(x = max_year,
    ymin = ci.lb,
    ymax = ci.ub,
    group = cvp_lab),alpha=0.1) +
  scale_color_manual(values=c("orange","darkred")) +  
  labs(x = "Final year of the study", y = "Effect size (Fisher's Z)", color = "Mean CVp") +
   new_scale_colour() +
geom_point(data = meta_FINAL_1y, aes(x = max_year, y = Z, col =meanCVp),  size = 2)  +
scale_colour_viridis() + 
theme_minimal()    + theme(aspect.ratio = (1))   

sup_fig12 

   ggsave(
  filename = "sup_fig12.pdf",
  plot     = sup_fig12,
  width    = 8,
  height   = 8,
  units    = "in"
)
    ggsave(
  filename = "sup_fig12.png",
  plot     = sup_fig12,
  width    = 8,
  height   = 8,
  units    = "in"
)
  
```




```{r plot with predicted data}
  
###### plot with interaction shown as three linear predictions
###### interaction terms as slopes
######
 
## choose values density for predictions
 
  newdat_lines2 <- expand.grid(
  max_year  = as.numeric(c(1980,2020)),
   meanCVp  = cvp_vals,
  KEEP.OUT.ATTRS = FALSE
)           
                 
 ## create prediction matrix 
  
X2 <- model.matrix(~ meanCVp * max_year, data = newdat_lines2)
colnames(X2)[colnames(X2) == "(Intercept)"] <- "intrcpt"              
                 
coef_names2 <- names(coef(modelCV))            # e.g. "intrcpt","cvp","time","cvp:time"
beta_names2 <- setdiff(coef_names2, "intrcpt")
X2 <- X2[, beta_names2, drop = FALSE]                
               
 pr2 <- predict(modelCV, newmods = X2)           
               
## prepare a data frame
pred_time_plot <- cbind(newdat_lines2,
                 pred  = as.numeric(pr2$pred),
                 ci.lb = as.numeric(pr2$ci.lb),
                 ci.ub = as.numeric(pr2$ci.ub))

## 6) Plot two lines (one per CVp) across time with CI ribbons

pred_time_plot$year_lab <- factor(round(pred_time_plot$max_year, 2))


ggplot(pred_time_plot, aes(x = meanCVp, y = pred, color = year_lab, fill = year_lab)) +
  geom_ribbon(aes(ymin = ci.lb, ymax = ci.ub), alpha = 0.15) +
  geom_line(linewidth = 1) +  
  scale_colour_manual(values=c("darkorange","darkblue")) +  
  labs(x = "CVp", y = "Effect size (Z)", color = "year", fill = "year") +
  theme_minimal()    

 
 SUP_CV <-  ggplot() + 
  geom_line(data=pred_time_plot, aes(x = meanCVp, y = pred, color = year_lab, fill = year_lab), linewidth = 1) +  
    geom_ribbon(data=pred_time_plot, aes(x = meanCVp,
    ymin = ci.lb,
    ymax = ci.ub,
    group = year_lab),alpha=0.1) +
  scale_color_manual(values=c("lightblue","darkorange")) +  
  labs(x = "Final year", y = "Effect size (Z)", color = "Final year") +
   new_scale_colour() +
  geom_point(data = meta_FINAL_1y, aes(x = meanCVp, y = Z,  size = 1/sqrt(var_z), col=max_year), alpha = 0.8)  +
scale_color_viridis_c() +
theme_minimal()  +    theme(aspect.ratio = (1))   

  SUP_CV 
  

  
  
```



```{r trends in time for hyptohesis 4 and within clades correlations}

# sample size and max year

ggplot() + geom_point(data = meta_FINAL_1y,aes(x = max_year,y = sample_size))

cor.test(meta_FINAL_1y$max_year,meta_FINAL_1y$sample_size)

ggplot() + geom_point(data = meta_FINAL_1y,aes(x = max_year, y = median_year))

### check the model with different plant types 

model_year_plant <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   max_year * plant_type,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y )
  
summary(model_year_plant)
summary(modelCV)
###  subset data to similar species Gymnosperms

modelCV_gymnosperms <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   meanCVp  * max_year,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y %>% filter(plant_type == "Gymnosperm"))
          
summary(modelCV_gymnosperms)

### subset data to similar species Angiosperms

modelCV_angiosperms <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   meanCVp  * max_year,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y %>% filter(plant_type == "Angiosperm"))
          
summary(modelCV_angiosperms)

#### Fagales
 
modelCV_fagales <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   meanCVp  * max_year,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id), 
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data =  meta_FINAL_1y %>% filter(genus == "Fagus" | genus == "Castanea" | genus == "Quercus") )
      
summary(modelCV_fagales)

##

modelCV_fagus <-  rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   max_year,
                       random = list(
                                    ~1|title,
                                      ~1|ES_id), 
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data =  meta_FINAL_1y %>% filter(genus == "Fagus" ) )
      
summary(modelCV_fagus)

####

 ggplot(data = meta_FINAL_1y %>% filter(plant_type == "Gymnosperm"), aes(meanCVp)) +
geom_histogram(col="black",fill = "darkgrey"  , bins = 25) + xlab("meanCVp")  
 
ggplot(data = meta_FINAL_1y %>% filter(plant_type == "Angiosperm"), aes(meanCVp)) +
geom_histogram(col="black",fill = "orange"  , bins = 25) + xlab("meanCVp")  

ggplot(data = meta_FINAL_1y, aes(meanCVp)) +
geom_histogram(col="black",fill = "orange"  , bins = 25) + xlab("meanCVp")  


####

## sensitivity analysis
## run rma.models on multiple data frames that reduce number of species or genus to see whether species turnover across time influences model results
## for each model report Estimate, Z-value and p-value
## draw histogram, distribution of the effects using all models

# Remove rows where species is missing
 
df_model <- meta_FINAL_1y %>%
  filter(
    !is.na(tip_label),
    !is.na(Z),
    !is.na(var_z),
    !is.na(meanCVp),
    !is.na(max_year),
    !is.na(tip_label),
    !is.na(title),
    !is.na(ES_id)
  ) %>%
  as.data.frame()

species_list <- unique(df_model$tip_label)

# Lists for models and extracted results
model_list  <- list()
result_list <- list()

for (i in seq_along(species_list)) {

  excluded_sp <- species_list[i]

  # Dataset containing all species except one
  df_subset <- df_model[
    df_model$tip_label != excluded_sp,
    ,
    drop = FALSE
  ]

  message(
    "Excluding: ", excluded_sp,
    "; rows remaining: ", nrow(df_subset),
    "; species remaining: ",
    length(unique(df_subset$tip_label))
  )
  # Run the same model on the reduced data frame
  model_i <- tryCatch(
    {
      metafor::rma.mv(
        yi = Z,
        V = var_z,
        mods = ~ meanCVp * max_year,
        random = list(
          ~ 1 | tip_label,
          ~ 1 | title,
          ~ 1 | ES_id
        ),
        R = list(
          tip_label = phylo_corMatrix
        ),
        method = "REML",
        control = list(
          optimizer = "optim",
          optmethod = "Nelder-Mead"
        ),
        data = df_subset
      )
    },
    error = function(e) {
      message(
        "Model failed after excluding ",
        excluded_sp,
        ": ",
        conditionMessage(e)
      )
      NULL
    }
  )
  
 # Continue to next species if the model failed
  
  if (is.null(model_i)) {
    next
  }
# Save the complete model object
  model_list[[as.character(excluded_sp)]] <- model_i

  # Save model coefficients
  result_list[[as.character(excluded_sp)]] <- data.frame(
    excluded_species = as.character(excluded_sp),
    term = rownames(model_i$b),
    estimate = as.numeric(model_i$b),
    se = as.numeric(model_i$se),
    z_value = as.numeric(model_i$zval),
    p_value = as.numeric(model_i$pval),
    ci_lower = as.numeric(model_i$ci.lb),
    ci_upper = as.numeric(model_i$ci.ub),
    k = model_i$k,
    n_species_remaining = dplyr::n_distinct(df_subset$species),
    stringsAsFactors = FALSE
  )
}

# Combine all results into one data frame

model_results <- bind_rows(result_list)
model_results_int  <- model_results %>% filter(term == "meanCVp:max_year")

## plot the data based on estimate, Z and p

modelCVp_hist1 <-  ggplot(data = model_results_int, aes(estimate)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) + xlab("estimate") + labs(title="A") +theme(aspect.ratio = (1))
 
modelCVp_hist2 <- ggplot(data = model_results_int, aes(z_value)) +
geom_histogram(col="black",fill = "blue"  , bins = 25) + xlab("z-value") + labs(title="B") +   theme(aspect.ratio = (1))

modelCVp_hist3 <- ggplot(data = model_results_int, aes(p_value)) +
geom_histogram(col="black",fill = "red"  , bins = 25) + xlab("p-value") + labs(title="C")  + theme(aspect.ratio = (1))

 ggplot(data = model_results_int, aes(p_value)) +
geom_histogram(col="black",fill = "red"  , bins = 25) + xlab("p-value") + labs(title="c") + xlim(0,0.005)

modelCV_hist <- modelCVp_hist1 + modelCVp_hist2 + modelCVp_hist3

modelCV_hist

############# THE SAME OPERATION FOR DIFFERENT GENERA 
 
# List of species that will be excluded one at a time
 
genus_list <- unique(genus_list$genus)

# Lists for models and extracted results
model_list_gen <- list()
result_list_gen <- list()

for (i in seq_along(genus_list)) {

  excluded_gen <- genus_list[i]

  # New data frame containing all species except excluded_sp
  df_subset <- df_model[
    df_model$genus != excluded_gen,
  ]

  # Remove incomplete observations required by the model
 
  message(
    "Excluding: ", excluded_gen,
    "; rows remaining: ", nrow(df_subset),
    "; genus remaining: ",
    length(unique(df_subset$genus))
  )

  # Run the same model on the reduced data frame
  model_i <- tryCatch(
    {
      rma.mv(
         yi = Z, V = var_z,
                        method = "REML",
                        mods = ~   meanCVp  * max_year,
                       random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
        data = df_subset
      )
    },
    error = function(e) {
      message(
        "Model failed after excluding ",
        excluded_gen,
        ": ",
        e$message
      )
      return(NULL)
    }
  )

  # Continue to next species if the model failed
  if (is.null(model_i)) {
    next
  }

  # Save the complete model object
  model_list_gen[[as.character(excluded_gen)]] <- model_i

  # Save model coefficients
  result_list_gen[[as.character(excluded_gen)]] <- data.frame(
    excluded_genus = as.character(excluded_gen),
    term = rownames(model_i$b),
    estimate = as.numeric(model_i$b),
    se = as.numeric(model_i$se),
    z_value = as.numeric(model_i$zval),
    p_value = as.numeric(model_i$pval),
    ci_lower = as.numeric(model_i$ci.lb),
    ci_upper = as.numeric(model_i$ci.ub),
    k = model_i$k,
    tau2 = model_i$tau2,
    n_genera_remaining = length(unique(df_subset$genus)),
    stringsAsFactors = FALSE
  )
}
 
# Combine all results into one data frame
model_results_gen <- bind_rows(result_list_gen)
model_results_gen_int  <- model_results_gen %>% filter(term == "meanCVp:max_year")

## plot the data based on estimate, Z and p
 modelCVp_hist4 <-  ggplot(data = model_results_gen_int, aes(estimate)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) + xlab("estimate") + labs(title="D") + theme(aspect.ratio = (1))
 
 modelCVp_hist5 <-  ggplot(data = model_results_gen_int, aes(z_value)) +
geom_histogram(col="black",fill = "blue"  , bins = 25) + xlab("z-value") + labs(title="E")  + theme(aspect.ratio = (1))

 modelCVp_hist6 <- ggplot(data = model_results_gen_int, aes(p_value)) +
geom_histogram(col="black",fill = "red"  , bins = 25) + xlab("p-value") + labs(title="F") + theme(aspect.ratio = (1))

 
 modelCV_hist_all <-   modelCV_hist + modelCVp_hist4 + modelCVp_hist5 + modelCVp_hist6

 ### save the FIG.S9
 
 modelCV_hist_all
 
ggsave(
  filename = "modelCV_hist_all.pdf",
  plot     = modelCV_hist_all,
#  device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 14,
  height   = 12,
  units    = "in"
)

```
 

```{r publication bias using entire data set}

### used in publication (FIG.S4)

model_PB <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ sqrt(var_z),
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL)

summary(model_PB)
 
## non-significant 

funnel(model_PB)

funnel(model_PB, 
      yaxis="seinv", # Inverse of standard error (precision) as the y axis
      level = c(90, 95),  # levels of statistical significance highlighted 
      shade = c("white", "gray55"), # shades for different levels of statistical significance
      legend = TRUE, # display legend
      ylab="Precision (1/SE)", 
      cex.lab=1.4, 
      digits=1)

# Heteroscedasticity check

qplot(y = sqrt(residuals(M4.1)^2), x = fitted(model_PB)) + # plot sqrt(residuals^2) against fitted values
      geom_point() +
      geom_smooth(method = "lm") + # method ="lm" to generate a straight line 
      geom_hline(yintercept = 0, colour="red") 
 

pred_Egger_reg4.1  <- orchaRd::mod_results(model_PB, mod = "var_z", group = "title",
                    weights = "prop")
                      
                    
orchaRd::bubble_plot(pred_Egger_reg4.1, group = "title",  mod = "var_z" , xlab = "Standard error", legend.pos = "top.left")  +
geom_point(data = meta_INC, aes(x = (var_z), y = Z, color = plant_type, size = 1/sqrt(var_z)), alpha = 0.6) + scale_color_discrete()

### plot_continuous(meta_FINAL, modelCV, sqrt(meta_FINAL$var_Z), "Standard error")
#####

model_time <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~ max_year,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL)

summary(model_time)

bubble_year2 <- orchaRd::mod_results(model_time, mod = "max_year", group = "title",
                    weights = "prop")
                    
orchaRd::bubble_plot(bubble_year2, group = "title",  mod = "max_year", xlab = "Year", legend.pos = "top.left") + 
geom_point(data = meta_FINAL, aes(x = max_year, y = Z, color = plant_type, size = 1/sqrt(var_z)), alpha = 0.6) + scale_color_discrete()
  
```




```{r PCAs of traits and climate}

#### running PCA with climatic variables
library(factoextra)

worldclim. <- na.omit(worldclim) 
worldclimPCA <- worldclim.[, !(names(worldclim.) %in% c("latitude", "longitude"))]
worldclimPCA <- prcomp(worldclimPCA, scale. = TRUE)

summary(worldclimPCA)
worldclimPCA$rotation

screeplot(worldclimPCA, type = "lines", main = "Scree Plot")
biplot(worldclimPCA, scale = 0)

#
fviz_pca_biplot(worldclimPCA)

pca_scores <- as.data.frame(worldclimPCA$x)
pca_scores$latitude <- worldclim.$latitude
pca_scores$longitude <- worldclim.$longitude

##

ggplot(pca_scores, aes(x = PC1, y = PC2, color = latitude)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA: PC1 vs PC2 colored by Latitude")
 

meta_FINAL_1y <- left_join(meta_FINAL_1y,pca_scores, by =c("longitude", "latitude"))
 

 


## running PCA with trait variables

SpeciesTraits <- meta_FINAL_1y %>%
dplyr::select(tip_label,plant_type,seed_mass, wood_density_imp, SLA, plant_height) %>% 
  rename("Seed mass" = seed_mass, "Plant height" = plant_height,"Wood density" = wood_density_imp) %>%
na.omit()  

SpeciesTraits <- SpeciesTraits[!duplicated(SpeciesTraits), ]

TraitsPCA <- prcomp(SpeciesTraits[,-1:-2], scale. = TRUE)

summary(TraitsPCA)
TraitsPCA$rotation

screeplot(TraitsPCA, type = "lines", main = "Scree Plot")
biplot(TraitsPCA, scale = 0)

 ### plot PCA with trait values

plot_traits_PCA <- fviz_pca_biplot(TraitsPCA, label = "var", col.var = "darkgreen" , habillage=SpeciesTraits$plant_type, geom = "point", pointsize = 3.5, alpha = 0.8,col.circle = "red",
               #addEllipses=TRUE, ellipse.level=0.95,
               arrowsize = 1, repel = TRUE, title = "A", mean.point = FALSE, 
               ggtheme = theme_minimal()) + scale_colour_manual(values=c("darkorange","black")) + 
               xlim(-4, 4) + ylim (-3, 3) + 
               theme(axis.title.x = element_text(size = 16), axis.title.y = element_text(size = 16), title = element_text(size=15)) + theme(aspect.ratio = (1)) +  my_theme + theme(legend.position = "bottom")  

plot_traits_PCA

########## ################################# #####################
################################# #####################
## plot histograms with trait values
    
hist(SpeciesTraits_hist_log$seed_mass, breaks = 20, col = "darkgreen", xlab = "Seed mass", main = "") 

hist(SpeciesTraits_hist_log$wood_density_imp, breaks = 20, col = "darkgreen" , xlab = "Wood density", main = "") 

 hist(SpeciesTraits_hist_log$plant_height, breaks = 20, col = "darkgreen" ,xlab = "Plant height", main = "") 

hist(SpeciesTraits_hist_log$SLA, breaks = 20, col = "darkgreen", xlab = "SLA", main = "") 

logSLA <-  ggplot(data = SpeciesTraits_hist_log, aes(SLA)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) + xlab("SLA") + labs(title="D")

logPH <- ggplot( data = SpeciesTraits_hist_log, aes(plant_height)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) +  xlab("Plant height" )+ labs(title="B")

logWD <-  ggplot( data = SpeciesTraits_hist_log, aes(wood_density_imp)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) +  xlab("Wood density")+ labs(title="E")
 
logSM <- ggplot( data = SpeciesTraits_hist_log, aes(seed_mass)) +
geom_histogram(col="black",fill = "green4"  , bins = 25) +  xlab("Seed mass")+ labs(title="C")
 
PCA_hist <- plot_traits_PCA + (logPH  + logSM +logSLA +logWD ) + my_theme

PCA_hist

## SAVE THE FIG.S10

ggsave(
  filename = "pca_hist_traits.pdf",
  plot     = PCA_hist,
  width    = 16,
  height   = 10,
  units    = "in"
)

########## ################################# #####################

pca_tr_scores <- as.data.frame(TraitsPCA$x)

pca_tr_scores$tip_label <- SpeciesTraits$tip_label

pca_tr_scores <- pca_tr_scores %>%dplyr::select(PC1,PC2,tip_label) %>% rename(pc1 = PC1, pc2 = PC2)
 
meta_FINAL_1y <- left_join(meta_FINAL_1y,pca_tr_scores, by = "tip_label") 


########## ################################# #####################

model_PCA_wc <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  PC1 + PC2,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_matrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(model_PCA_wc)    

model_PCA_traits <- rma.mv(yi = Z, V = var_z,
                        method = "REML",
                        mods = ~  pc1 + pc2,
                        random = list(~1|tip_label,
                                      ~1|title,
                                      ~1|ES_id),
                        R =  list(tip_label = phylo_corMatrix),
                        control=list(optimizer="optim",optmethod="Nelder-Mead"),
                        data = meta_FINAL_1y)
      
summary(model_PCA_traits)    
                      
################### no effects of traits and climate

```

 
```{r plotting a map}

# for mapping geoms 
library(rnaturalearth)
# for map data 
library(ggtext)

world_coordinates <- map_data("world")
 
#create world map using ggplot() function
ggplot() + geom_map( data = world_coordinates, map = world_coordinates, aes(x=long, y=lat, map_id = region), fill ="darkgreen", alpha = 0.8) +
geom_point(data=meta_FINAL, aes(x=longitude, y=latitude), shape = "circle", col = "darkorange", size = 2) + theme_void()

## remove antarctica
world_coordinates2 <- subset(world_coordinates, world_coordinates$region != "Antarctica")   

world_coordinates2 <- subset(world_coordinates, long > -170   & lat > -55 )
 
# as input to plot world map

 plotMapLeg <- ggplot() + geom_map( data = world_coordinates2, map = world_coordinates2, aes(x=long, y=lat, map_id = region), fill ="darkgreen", alpha = 0.8) +
 geom_point( data=meta_FINAL, aes(x=longitude, y=latitude, group = genus, col = plant_type), shape = "circle", size = 2.5) +
theme_void() + scale_color_manual(values=c("darkorange","black")) + theme(legend.title = element_text(size=24), legend.text = element_text(size=20)) +   labs(color = "Plant type")  

 leg <- ggpubr::get_legend(plotMapLeg)

ggsave(
  filename = "leg.pdf",
  plot     = leg,
#  device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 14,
  height   = 12,
  units    = "in"
)

# coord_sf()

plotMap1 <-  ggplot() + geom_map( data = world_coordinates2, map = world_coordinates2, aes(x=long, y=lat, map_id = region),
fill ="darkgreen", alpha = 0.8) +
 geom_point( data=meta_FINAL, aes(x=longitude, y=latitude, group = genus, col = plant_type), 
 shape = "circle", size = 2.5 ,colour = "gray95",  stroke = 1) +
  geom_point( data=meta_FINAL, aes(x=longitude, y=latitude, group = genus, col = plant_type), shape = "circle", size = 2) +  labs(title = "A") +
theme_void() + scale_color_manual(values=c("darkorange","black")) + 
theme(legend.text = element_text(size=15), plot.title  = element_text(size=25),legend.title = element_text(size = 20)) + 
  theme(legend.position = c(.1, .15)) + labs(color = "Plant type")
# theme(legend.position =  "none")  

plotMap1

plotMapLeg

 ggsave(
  filename = "plotMapSOLO.png",
  plot     = plotMapLeg,
#  device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 20,
  height   = 10,
  units    = "in"
)

```


PLOTTING FIG.1

```{r plot biomes}

meta_FINAL_1y$MAT = meta_nonsc_trait_1y$`Annual Mean Temperature`

meta_FINAL_1y$MAP = meta_nonsc_trait_1y$`Annual Precipitation`/10


ggplot() + geom_point(data = meta_FINAL_1y,aes(x = MAT,y = MAP))  

ggplot() + geom_point(data = meta_FINAL_1y,aes(x = as.numeric(rainfall),y = MAP))  

## plot all data
whittaker_base_plot() +
  geom_point(data = meta_FINAL_1y,
             aes(x = MAT,
                 y = MAP),
                 colour = "black", alpha = 0.5,
             shape = 16,
             size  = 4) +
  theme_bw()
   
## plot for manuscript
 plot1c <- whittaker_base_plot() +
   geom_point(data = meta_FINAL_1y,
             aes(x = MAT,
                 y = MAP),
             shape  = 21,
             stroke = 1, # acts as the thickness of the boundary line
             colour = "gray95", # acts as the color of the boundary line
             size   = 3.5) +
               geom_point(data = meta_FINAL_1y,
             aes(x = MAT,
                 y = MAP,
                 colour = plant_type), # map `status` variable to colour
             shape = 16,
             size  = 3,
             alpha = 0.8) +
             scale_colour_manual(values=c("darkorange","black")) +
   labs(title = "C") +    ylab("MAP [cm]") +
    xlab("MAT [°C]") + my_theme +
  theme_bw() +  theme(axis.title = element_text(size = 20),
                      axis.text = element_text(size = 18),
                      plot.title  = element_text(size=24)) +  theme(aspect.ratio = (1)) +
theme(legend.position = "none")
  
plot1c

### plot of climate, map and pca to supplement
# plotMap1  / (plot_traits_PCA + plotMap2 )

## plot with data description

  plotMap1  / (plot1b + plot1c ) +  plot_layout(guides='collect') + 
    inset_element(leg, 0,1,0.5,1, align_to = "full", clip=TRUE) + my_theme
 
 
plotMap <- plotMap1  / (plot1b + plot1c ) + 
  plot_layout(widths = c(1, 2)) 

### FIG.1

plotMap 

## 3) Save to a high-quality PDF (vector, embedded fonts)
# Pick a sensible width/height; ~16x7 inches works well for two side-by-side panels.
# Use device = cairo_pdf for robust font embedding across platforms.

ggsave(
  filename = "plotMap-noLeg.pdf",
  plot     = plotMap,
#  device   = cairo_pdf,   # needs Cairo installed; on Windows it ships with R, on macOS use XQuartz if needed
  width    = 14,
  height   = 12,
  units    = "in"
)

```


### FIGURE 2 (PHYLOGENY) can be found in additional code
 

```{r pressure, echo=FALSE}

## show information about session
sessionInfo()

## show how to cite packages
citation(package='metafor')
citation(package='plotbiomes')
citation(package='ape')
citation(package='orchaRd')

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


